MISUSES AND INACCURACIES OF PAGEL'S LAMBDA ML ESTIMATION METHODS

### 2020-04-24 Manuscript/Manuscript-2.Rmd
vsn 2

The results above demonstate that $\lambda$ is not a reliable estimate of the phylogenetic signal in phenotypic data. As such, biological interpretations of the strength of phylogenetic signal, and comparisons of the magnitude of such effects across datasets, are severely compromised when based on this parameter. As an alternative, we propose that summary estimates of phylogenetic signal be converted to effect sizes for these purposes. A standardized effect size is found as:

@@

where $\theta_{obs}$ is the observed test statistic, $E(\theta)$ is its expected value under the null hypothesis, and $\sigma_\theta$ is its standard error [@Glass1976; @Cohen1988;@Rosenthal1994]. $Z_{\theta}$ expresses the magnitude of the effect in $\theta_{obs}$ by transforming the original test statistic to a standard normal deviate [@Glass1976; @Kelley2012]. Here, $\theta_{obs}$ and $\sigma_\theta$ are estimated from the data, while $E(\theta)$ is obtained from the distribution of $\theta$ derived from parametric theory. However, recent advances in resampling theory [@Collyer_et_al2015a;@AdamsCollyer2016; @AdamsCollyer2019b] have shown that $E(\theta)$ and $\sigma_\theta$ may also be obtained from an empirical sampling distribution obtained from permutation procedures. \hfill\break

@@

Note that under the null hypothesis, $E(\lambda)=0$, a no phylogenetic signal is expected under this condition [@Freckleton_et_al2002].

  - Z-score. could be Lambda or Kappa.  Show it is Kappa
-Comparing the strength of physig
  - two sample Z-score

-Conclusions and Implications

Finally, for comparison we characterized the strength of phylogenetic signal in each dataset using a standardized effect size [$Z_K$: sensu @AdamsCollyer2016; @AdamsCollyer2019b] based on $Kappa$.

Variation in the set of $Z_K$ at each input level of phylogenetic signal was then calculated as an estimate of precision in $Z_K$. However, because $Z_K$ differs in scale from $\lambda$, we used a linear normalization to standardize $Z_K$ to a uniform distribution ($0\rightarrow1$), and estimated the precision of $Z_K$ from the normalized values. \hfill\break

## *Literature Survey*

@@

# Results

## *Simulations*

By contrast, when characterizing phylogenetic signal with the standardized effect size ($Z_K$) of $Kappa$, we found that the precision of $Z_K$ was more stable, as variation across datasets was far more consistent across the range of input values. For example, when $n=128$ the precision of $\lambda_{est}$ (Figure 2A) varied considerably more across input levels of phylogenetic signal than did the precision of $Z_K$ for the same datasets (Figure 2B). Further, for a given $n$, the variance in precision estimates of $Z_K$ was considerably smaller than the variance in precision estimates of $\lambda_{est}$ (Fig. 2C,D); implying that estimates of phylogenetic signal were more reliable and robust when using $Z_K$ (for additional results see Supporting Information).  Finally, it should also be recognized that because $Z_K$ is a standardized effect size, the strength of phylogenetic signal is more readily interpretable when using this measure, as $Z_K$ expresses the strength of phylogenetic signal in standard deviation units relative to the mean [see @AdamsCollyer2019]. This further implies that comparisons of the strength of phylogenetic signal among phenotypic traits are possible, and may be accomplished statistically via a two-sample test that formally compares $Z_K$ across datasets [for comparisons of multivariate effect sizes see: @AdamsCollyer2016; @AdamsCollyer2019b]. \hfill\break

## *Literature Survey*

We found 182 manuscripts from 2019 that estimated and reported Pagel's lambda values using PGLS methods. These papers averaged 8.527 lambda values, ranging from a single lambda estimate up to 71 estimated lambdas. Almost exactly half of the published lambda estimates were either below 0.05 (25.32%) or above 0.9 (24.74%; Figure 3). 73.32% of the published lambdas were estimated using phylogenies with fewer than 200 tips, and 348 lambda estimates (8.57% of all published estimates) came from phylogenies with fewer than 30 tips.  \hfill\break

@@

# Results

>>>>

The results above demonstate that $\lambda$ is not a reliable estimate of the phylogenetic signal in phenotypic data. As such, biological interpretations of the strength of phylogenetic signal, and comparisons of the magnitude of such effects across datasets, are severely compromised when based on this parameter. As an alternative, we propose that summary estimates of phylogenetic signal be converted to effect sizes for interpretation and comparison. Statistically, a standardized effect size may be found as:

@@

where $\theta_{obs}$ is the observed test statistic, $E(\theta)$ is its expected value under the null hypothesis, and $\sigma_\theta$ is its standard error [@Glass1976; @Cohen1988;@Rosenthal1994]. $Z_{\theta}$ expresses the magnitude of the effect in $\theta_{obs}$ by transforming the original test statistic to a standard normal deviate [@Glass1976; @Kelley2012]. Typically, $\theta_{obs}$ and $\sigma_\theta$ are estimated from the data, while $E(\theta)$ is obtained from the distribution of $\theta$ derived from parametric theory. However, recent advances in resampling theory [@Collyer_et_al2015a;@AdamsCollyer2016; @AdamsCollyer2019b] have shown that $E(\theta)$ and $\sigma_\theta$ may also be obtained from an empirical sampling distribution of $\theta$ obtained from permutation procedures. \hfill\break

@@

Note that under the null hypothesis, $E(\lambda)=0$, a no phylogenetic signal is expected under this condition [@Freckleton_et_al2002]. \hfill\break

To evaluate the utility of $Z_K$ and $Z_\lambda$ we calculated both effect sizes for the simulated datasets generated above, and summarized the precision of each using its variance ($\sigma^2_{Z_K}$ and $\sigma^2_{Z_\lambda}$). Results are found in Figure 3. Here two things are evident. First, estimates of $Z_K$ track the input phylogenetic signal in a more linear fashion than to estimates of $Z_\lambda$. Second, the precision of $Z_K$ is considerably more stable, as coefficients of variation for the set of $\sigma^2_{Z_K}$ across input levels of phylogenetic signal for the same sample size was an order of magnitude smaller than that observed for $\sigma^2_{Z_\lambda}$ (Figure 3). This implies that estimates of the strength of phylogenetic signal were more reliable and robust when using $Z_K$ as compared with $Z_\lambda$. \hfill\break

[insert Figure 3 here]

## *Statistical Comparisons of Phylogenetic Signal*

Once the magnitude of phylogenetic signal is characterized using $Z_K$, it may be of interest to compare such measures across datasets. This is useful, for instance, to determine whether the strength of phylogenetic signal is greater in one phenotypic trait as compared with another. As with other effect sizes derived from permutation distributions [e.g., @AdamsCollyer2016; @AdamsCollyer2019b], a two-sample test statistic may be found as:

\begin{align}
  \hat{Z}_{12}=\frac{\lvert{(K_{1}-\hat\mu_{K_1})-(K_{2}-\hat\mu_{K_2})}\rvert}{\sqrt{\hat\sigma^2_{K_1}+\hat\sigma^2_{K_2}}}
\end{align}

where $K_1$, $K_2$, $\hat\mu_{K_1}$, $\hat\mu_{K_2}$, $\hat\sigma_{K_1}$, and $\hat\sigma_{K_2}$ are as defined above for equation 2. Estimates of significance of $\hat{Z}_{12}$ may be obtained from a standard normal distribution. As with other two-sample tests, $\hat{Z}_{12}$ is typically considered a two-tailed test, however directional (one-tailed) tests may be specified should the empirical situation require it [see @AdamsCollyer2016; @AdamsCollyer2019b].

## *Empirical Example*

# Conclusions and Implications

## *Literature Survey... WHERE DOES THIS GO???*

@@

*Results*. We found 182 manuscripts from 2019 that estimated and reported Pagel's lambda values using PGLS methods. These papers averaged 8.527 lambda values, ranging from a single lambda estimate up to 71 estimated lambdas. Almost exactly half of the published lambda estimates were either below 0.05 (25.32%) or above 0.9 (24.74%; Figure 3). 73.32% of the published lambdas were estimated using phylogenies with fewer than 200 tips, and 348 lambda estimates (8.57% of all published estimates) came from phylogenies with fewer than 30 tips.  \hfill\break

@@

### 2020-04-24 Manuscript/Manuscript-2.tex
vsn 2

The results above demonstate that \(\lambda\) is not a reliable estimate
of the phylogenetic signal in phenotypic data. As such, biological
interpretations of the strength of phylogenetic signal, and comparisons
of the magnitude of such effects across datasets, are severely
compromised when based on this parameter. As an alternative, we propose
that summary estimates of phylogenetic signal be converted to effect
sizes for these purposes. A standardized effect size is found as:

@@

where \(\theta_{obs}\) is the observed test statistic, \(E(\theta)\) is
its expected value under the null hypothesis, and \(\sigma_\theta\) is
its standard error (Glass 1976; Cohen 1988; Rosenthal 1994).
\(Z_{\theta}\) expresses the magnitude of the effect in \(\theta_{obs}\)
by transforming the original test statistic to a standard normal deviate
(Glass 1976; Kelley and Preacher 2012). Here, \(\theta_{obs}\) and
\(\sigma_\theta\) are estimated from the data, while \(E(\theta)\) is
obtained from the distribution of \(\theta\) derived from parametric
theory. However, recent advances in resampling theory (Collyer et al.
2015; Adams and Collyer 2016, 2019a) have shown that \(E(\theta)\) and
\(\sigma_\theta\) may also be obtained from an empirical sampling
distribution obtained from permutation procedures. \hfill\break

@@

Note that under the null hypothesis, \(E(\lambda)=0\), a no phylogenetic
signal is expected under this condition (Freckleton et al. 2002).

\begin{itemize}
\tightlist
\item
  Z-score. could be Lambda or Kappa. Show it is Kappa -Comparing the
  strength of physig
\item
  two sample Z-score
\end{itemize}

-Conclusions and Implications

Finally, for comparison we characterized the strength of phylogenetic
signal in each dataset using a standardized effect size (\(Z_K\): sensu
Adams and Collyer 2016, 2019a) based on \(Kappa\).

Variation in the set of \(Z_K\) at each input level of phylogenetic
signal was then calculated as an estimate of precision in \(Z_K\).
However, because \(Z_K\) differs in scale from \(\lambda\), we used a
linear normalization to standardize \(Z_K\) to a uniform distribution
(\(0\rightarrow1\)), and estimated the precision of \(Z_K\) from the
normalized values. \hfill\break

\hypertarget{literature-survey}{%
\subsection{\texorpdfstring{\emph{Literature
Survey}}{Literature Survey}}\label{literature-survey}}

@@

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{simulations}{%
\subsection{\texorpdfstring{\emph{Simulations}}{Simulations}}\label{simulations}}

By contrast, when characterizing phylogenetic signal with the
standardized effect size (\(Z_K\)) of \(Kappa\), we found that the
precision of \(Z_K\) was more stable, as variation across datasets was
far more consistent across the range of input values. For example, when
\(n=128\) the precision of \(\lambda_{est}\) (Figure 2A) varied
considerably more across input levels of phylogenetic signal than did
the precision of \(Z_K\) for the same datasets (Figure 2B). Further, for
a given \(n\), the variance in precision estimates of \(Z_K\) was
considerably smaller than the variance in precision estimates of
\(\lambda_{est}\) (Fig. 2C,D); implying that estimates of phylogenetic
signal were more reliable and robust when using \(Z_K\) (for additional
results see Supporting Information). Finally, it should also be
recognized that because \(Z_K\) is a standardized effect size, the
strength of phylogenetic signal is more readily interpretable when using
this measure, as \(Z_K\) expresses the strength of phylogenetic signal
in standard deviation units relative to the mean (see Adams and Collyer
2019b). This further implies that comparisons of the strength of
phylogenetic signal among phenotypic traits are possible, and may be
accomplished statistically via a two-sample test that formally compares
\(Z_K\) across datasets (for comparisons of multivariate effect sizes
see: Adams and Collyer 2016, 2019a). \hfill\break

\hypertarget{literature-survey-1}{%
\subsection{\texorpdfstring{\emph{Literature
Survey}}{Literature Survey}}\label{literature-survey-1}}

We found 182 manuscripts from 2019 that estimated and reported Pagel's
lambda values using PGLS methods. These papers averaged 8.527 lambda
values, ranging from a single lambda estimate up to 71 estimated
lambdas. Almost exactly half of the published lambda estimates were
either below 0.05 (25.32\%) or above 0.9 (24.74\%; Figure 3). 73.32\% of
the published lambdas were estimated using phylogenies with fewer than
200 tips, and 348 lambda estimates (8.57\% of all published estimates)
came from phylogenies with fewer than 30 tips. \hfill\break

@@

\hypertarget{results-1}{%
\section{Results}\label{results-1}}

>>>>

The results above demonstate that \(\lambda\) is not a reliable estimate
of the phylogenetic signal in phenotypic data. As such, biological
interpretations of the strength of phylogenetic signal, and comparisons
of the magnitude of such effects across datasets, are severely
compromised when based on this parameter. As an alternative, we propose
that summary estimates of phylogenetic signal be converted to effect
sizes for interpretation and comparison. Statistically, a standardized
effect size may be found as:

@@

where \(\theta_{obs}\) is the observed test statistic, \(E(\theta)\) is
its expected value under the null hypothesis, and \(\sigma_\theta\) is
its standard error (Glass 1976; Cohen 1988; Rosenthal 1994).
\(Z_{\theta}\) expresses the magnitude of the effect in \(\theta_{obs}\)
by transforming the original test statistic to a standard normal deviate
(Glass 1976; Kelley and Preacher 2012). Typically, \(\theta_{obs}\) and
\(\sigma_\theta\) are estimated from the data, while \(E(\theta)\) is
obtained from the distribution of \(\theta\) derived from parametric
theory. However, recent advances in resampling theory (Collyer et al.
2015; Adams and Collyer 2016, 2019a) have shown that \(E(\theta)\) and
\(\sigma_\theta\) may also be obtained from an empirical sampling
distribution of \(\theta\) obtained from permutation procedures.
\hfill\break

@@

Note that under the null hypothesis, \(E(\lambda)=0\), a no phylogenetic
signal is expected under this condition (Freckleton et al. 2002).
\hfill\break

To evaluate the utility of \(Z_K\) and \(Z_\lambda\) we calculated both
effect sizes for the simulated datasets generated above, and summarized
the precision of each using its variance (\(\sigma^2_{Z_K}\) and
\(\sigma^2_{Z_\lambda}\)). Results are found in Figure 3. Here two
things are evident. First, estimates of \(Z_K\) track the input
phylogenetic signal in a more linear fashion than to estimates of
\(Z_\lambda\). Second, the precision of \(Z_K\) is considerably more
stable, as coefficients of variation for the set of \(\sigma^2_{Z_K}\)
across input levels of phylogenetic signal for the same sample size was
an order of magnitude smaller than that observed for
\(\sigma^2_{Z_\lambda}\) (Figure 3). This implies that estimates of the
strength of phylogenetic signal were more reliable and robust when using
\(Z_K\) as compared with \(Z_\lambda\). \hfill\break

{[}insert Figure 3 here{]}

\hypertarget{statistical-comparisons-of-phylogenetic-signal}{%
\subsection{\texorpdfstring{\emph{Statistical Comparisons of
Phylogenetic
Signal}}{Statistical Comparisons of Phylogenetic Signal}}\label{statistical-comparisons-of-phylogenetic-signal}}

Once the magnitude of phylogenetic signal is characterized using
\(Z_K\), it may be of interest to compare such measures across datasets.
This is useful, for instance, to determine whether the strength of
phylogenetic signal is greater in one phenotypic trait as compared with
another. As with other effect sizes derived from permutation
distributions (e.g., Adams and Collyer 2016, 2019a), a two-sample test
statistic may be found as:

\begin{align}
  \hat{Z}_{12}=\frac{\lvert{(K_{1}-\hat\mu_{K_1})-(K_{2}-\hat\mu_{K_2})}\rvert}{\sqrt{\hat\sigma^2_{K_1}+\hat\sigma^2_{K_2}}}
\end{align}

where \(K_1\), \(K_2\), \(\hat\mu_{K_1}\), \(\hat\mu_{K_2}\),
\(\hat\sigma_{K_1}\), and \(\hat\sigma_{K_2}\) are as defined above for
equation 2. Estimates of significance of \(\hat{Z}_{12}\) may be
obtained from a standard normal distribution. As with other two-sample
tests, \(\hat{Z}_{12}\) is typically considered a two-tailed test,
however directional (one-tailed) tests may be specified should the
empirical situation require it (see Adams and Collyer 2016, 2019a).

\hypertarget{empirical-example}{%
\subsection{\texorpdfstring{\emph{Empirical
Example}}{Empirical Example}}\label{empirical-example}}

\hypertarget{conclusions-and-implications}{%
\section{Conclusions and
Implications}\label{conclusions-and-implications}}

\hypertarget{literature-survey-where-does-this-go}{%
\subsection{\texorpdfstring{\emph{Literature Survey\ldots{} WHERE DOES
THIS
GO???}}{Literature Survey\ldots{} WHERE DOES THIS GO???}}\label{literature-survey-where-does-this-go}}

@@

\emph{Results}. We found 182 manuscripts from 2019 that estimated and
reported Pagel's lambda values using PGLS methods. These papers averaged
8.527 lambda values, ranging from a single lambda estimate up to 71
estimated lambdas. Almost exactly half of the published lambda estimates
were either below 0.05 (25.32\%) or above 0.9 (24.74\%; Figure 3).
73.32\% of the published lambdas were estimated using phylogenies with
fewer than 200 tips, and 348 lambda estimates (8.57\% of all published
estimates) came from phylogenies with fewer than 30 tips. \hfill\break

@@
